Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-19 paper.qmd
Updated discussion

The effectiveness of this approach is illustrated through various examples. For instance, the two-curvy clusters example demonstrates how the model accurately fits the points, capturing both local and global structures in high-dimensional space. Our simulation case study further, five Gaussian cluster example shows that while all observed NLDR methods preserve the global structure, only tSNE effectively maintains the local structure, highlighting the specific strengths and quirks of different methods.

@@

In conclusion, while our method effectively captures and represents high-dimensional data structures, further enhancements could involve introducing approaches to bind the data, indicate line segments beyond \gD{}, and diagnose the fitted model. These improvements would help in creating a more accurate representation of the data when \gD{} layout is inadequate.

>>>>

With the model fit, how the NLDR methods organize points in \gD{} layout is examined using the simulated $4\text{-}D$ data having five Gaussian clusters. tSNE, UMAP, and PaCMAP preserved the global structure having well-separated clusters, while tSNE, and UMAP effectively maintains the local structure, showing the fitted model as *crumpled \gD{} sheet* that stretches in all four dimensions. The interesting result is that the model fit of PaCMAP show *pancake* shape.

@@

To extend layouts beyond $k$-D, when 2D is clearly inadequate, can use k-means instead of bin centroids.

In conclusion, while our method effectively captures and represents high-dimensional data structures, further enhancements could involve introducing approaches to diagnose the fitted model. These improvements would help in creating a more accurate representation of the data when \gD{} layout is inadequate.
